---
title: "MAS 812: Non-parametric methods"
author: "Gift Pamba"
date: "2026-04-29"
output:
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(DescTools)
library(kableExtra)

```

# ONE

A researcher collects the following data set representing the body mass index (BMI) measurements of participants in a health study

|      |      |      |      |      |      |      |      |      |      |      |      |      |      |      |
|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| 22.5 | 24.1 | 28.3 | 26.7 | 23.8 | 31.2 | 29.5 | 25.4 | 27.9 | 33.6 | 24.8 | 30.1 | 26.3 | 21.9 | 34.2 |

## A

Compute the **order statistics** for this data set.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Order observations in ascending order
bmi <- c(22.5, 24.1, 28.3, 26.7, 23.8, 
         31.2, 29.5, 25.4, 27.9, 33.6,
         24.8, 30.1, 26.3, 21.9, 34.2)

cat("Order statistics: ","\n",
  paste0(
    sort(x = bmi, decreasing = FALSE)
  )
)
```

## B

Find the **median**, **first quartile**(Q1), and **third quartile**(Q3), using the order statistics

```{r, echo = FALSE, warning = FALSE, message = FALSE}

q1 <- quantile(bmi, probs = 0.25)
q3 <- quantile(bmi, probs = 0.75)


cat(
  paste0("Median: ", median(bmi), "\n",
         "First quartile: ", q1, "\n",
         "Third quartile: ", q3)
)

```

## C

Calculate the **interquartile range(IQR)** and identify any potential outliers using the 1.5\*IQR rule

```{r, echo = FALSE, message = FALSE, warning = FALSE}

iqr <- IQR(bmi)

cat(
  paste0("Interquartile range (IQR): ", iqr, "\n", "\n", "outliers:", "\n")
)

# outlier lower bound
l <- q1 - 1.5*iqr

# outlier upper bound
q <- q3 + 1.5*iqr

# Outlier detection
outliers <- bmi[bmi>=q | bmi<=l]

if(length(outliers) > 0){
  cat(paste(outliers))
}else{
  cat(paste0("There are no outliers!"))
}



# cat(
#   paste0("[", l, ", ", q, "]")
#   )

```

# TWO

A researcher wants to test whether the median waiting time(in minutes) at a clinic is less than 20 minutes. A sample of 12 patients yields the following waiting times:

|     |     |     |     |     |     |     |     |     |     |     |     |
|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| 18  | 22  | 15  | 19  | 25  | 16  | 21  | 14  | 17  | 23  | 24  | 13  |

```{r, echo = FALSE, message = FALSE, warning = FALSE}
clinic_waiting <- c(18,22,15,19,25,16,21,14,17,23,24,13)
```

## A

Perform the **sign test** at $\alpha = 0.05$. State your hypotheses clearly.

## B

Perform the **Wilcoxon-signed rank test** at $\alpha = 0.05$.

## C

Compare the results of the two tests. Which test is more appropriate for these data and why?

# THREE

A study compares the effectiveness of two different diets on weight loss (in kg). The results are:

|            |     |     |     |     |     |     |     |     |     |
|------------|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| **Diet A** | 4.2 | 5.1 | 3.8 | 6.0 | 4.5 | 5.5 | 4.9 | 3.5 |     |
| **Diet B** | 2.5 | 3.2 | 2.8 | 4.0 | 3.0 | 3.5 | 2.2 | 2.9 | 3.1 |

```{r, echo = FALSE, warning = FALSE, message = FALSE}
weightloss <- data.frame(
  diet = c(rep("A",8), rep("B",9)),
  weight = c(4.2,5.1,3.8,6,4.5,5.5,4.9,3.5,
           2.5,3.2,2.8,4,3,3.5,2.2,2.9,3.1
  )
)

```

## A

Perform a **Wilcoxon rank-sum test** at $\alpha = 0.05$ to determine if the diets differ in weight loss.

## B

Perform a **median test** (Mood's median test) at $\alpha = 0.05$ for the same data.

## C

Perform a **Kolmogorov-Smirnov two-sample test** at $\alpha = 0.05$.

# FOUR

A researcher compares the effectiveness of three different teaching methods on test scores. Students are randomly assigned to one of the three methods, and their scores are recorded.

|              |     |     |     |     |     |     |     |     |
|--------------|-----|-----|-----|-----|-----|-----|-----|-----|
| **Method A** | 78  | 82  | 75  | 80  | 77  | 81  | 79  | 76  |
| **Method B** | 68  | 72  | 65  | 70  | 67  | 71  | 69  | 66  |
| **Method C** | 85  | 88  | 82  | 87  | 84  | 86  | 83  | 81  |

```{r, echo = FALSE, message = FALSE, warning = FALSE}
teaching <- data.frame(
  method = rep(c("A","B","C"), each = 8),
  score = c(78,82,75,80,77,81,79,76,
           68,72,65,70,67,71,69,66,
           85,88,82,87,84,86,83,81)
)

```

## A

Perform a **Kruskal-Wallis test** at $\alpha = 0.05$ to determine if the teaching methods differ

```{r, echo=FALSE, message=FALSE, warning=FALSE}

tm <- kruskal.test(score~method, data = teaching)

cat(
  paste0("Kruskal Wallis test", "\n",
         "p-value: ", round(tm$p.value,4), "\n",
         "Decision: ", ifelse(tm$p.value<0.05, yes = "Reject null hypothesis", no = "Fail 
                              to reject the null hypothesis"))
)
```

The sample data provides sufficient evidence against the null hypothesis that states that all three teaching methods share the same location in terms of student performance.

## B

Conduct **post-hoc pairwise comparisons** using the Dunn's test with Bonferroni correction and identify which methods differ significantly

```{r, echo = FALSE, message = FALSE, warning = FALSE}
dunn_teaching <- DunnTest(x = teaching$score, g = teaching$method,
                          method = "bonferroni",
                          alternative = "two.sided")

dunn_teaching |> as.matrix()


```

Fronm the results table above, teaching Method C and B has the only p-value less than 0.05. This means that both the teaching methods significantly differ from each other at a significance level of 5%

# FIVE

## A

A study measuring the reaction times (in milliseconds) for 6 subjects under three different lighting conditions in a repeated measures design:

|                  |     |     |     |     |     |     |
|------------------|-----|-----|-----|-----|-----|-----|
| **Subject**      | 1   | 2   | 3   | 4   | 5   | 6   |
| **Dim light**    | 250 | 265 | 258 | 272 | 248 | 260 |
| **Normal light** | 235 | 248 | 240 | 255 | 232 | 245 |
| **Bright light** | 220 | 232 | 225 | 238 | 218 | 228 |

```{r, echo = FALSE, message = FALSE, warning = FALSE}
light <- data.frame(
  subject = rep(1:6, each = 3),
  intensity = rep(c("dim", "normal", "bright"), 3),
  time = c(250, 235, 220, 265, 248, 232,
           258, 240, 225, 272, 255, 238,
           248, 232, 218, 260, 245, 228)
)
```

### a

Perform a **Friedman test** at $\alpha = 0.05$ to determine if the lighting conditions affect reaction time

```{r, echo = FALSE, message = FALSE, warning = FALSE}
lt <- friedman.test(y = light$time, groups = light$intensity, blocks = light$subject)

cat(
  paste0("Friedman test", "\n",
         "p-value: ",round(lt$p.value, 4), "\n",
         "Decision: ", ifelse(lt$p.value<0.05, 
                             yes = "Reject the null hypothesis",
                             no = "Fail to reject the null hypothesis"))
)
```

The sampled obervations show that that the reaction times under the three lighting conditions differ significantly with a 5% margin of error.

### b

Perform a post-hoc comparison using **Wilcoxon signed-rank tests** with an appropriate multiple comparison correction technique

```{r, echo = FALSE, message = FALSE, warning = FALSE}
light_wsr <- pairwise.wilcox.test(light$time, light$intensity,
                                  paired = TRUE, p.adjust.method = "bonferroni")
light_wsr$p.value

```

## B

A study examines the relationship between daily coffee consumption (cups per day) and hours of sleep per night for 10 individuals

|                |     |     |     |     |     |     |     |     |     |     |
|----------------|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| **individual** | 1   | 2   | 3   | 4   | 5   | 6   | 7   | 8   | 9   | 10  |
| **cups**       | 3   | 1   | 4   | 2   | 5   | 3   | 6   | 2   | 4   | 1   |
| **hours**      | 7.5 | 8.2 | 6.8 | 7.9 | 6.2 | 7.4 | 5.8 | 8   | 6.5 | 7.8 |

### a

Compute **Spearman's rank correlation coefficient** $\rho$ for these data

```{r, echo = FALSE, message = FALSE, warning = FALSE}
coffee <- data.frame(
  "individual" = seq(1,10,1),
  "cups" = c(3,1,4,2,5,3,6,2,4,1),
  "hrs" = c(7.5,8.2,6.8,7.9,6.2,7.4,5.8,8.0,6.5,7.8)
)

coffee_cor <- cor.test(
  x = coffee$cups, y = coffee$hrs,
  method = "spearman", alternative = "two.sided"
)

cat(
  paste0("Spearman's coefficient: ", round(coffee_cor$estimate, 4))
)
```

### b

Test whether $\rho$ is significantly different from zero at $\alpha = 0.05$. State your hypotheses clearly

$$
{H}_{0}: Relationship\ between\ cups\ consumed\ and\ hours\ slept\ is\ non-monotonic \\ {H}_{1}: Relationship\ between\ cups\ consumed\ and\ hours\ slept\ is\ monotonic 
$$

```{r, echo = FALSE, message = FALSE, warning = FALSE}
cat(
  paste0("Spearman rank coefficient:", "\n",
         "p-value: ", round(coffee_cor$p.value, 4), "\n",
         "Decision: ", ifelse(coffee_cor$p.value<0.05,
                             yes = "Reject the null hypothesis",
                             no = "Fail to reject the null hypothesis"))
)
```

In conclusion, there is sufficient evidence in favor of the claim that the relationship between the number of coffee cups consumed in a day and the number of sleep hours at night is monotonic in nature at a significance level of 5%
