---
title: "MAS 812:Non-Parametric Methods"
author: "Gift Pamba"
date: "MSC/MAT/00098/024"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(rmarkdown)
library(tidyverse)
library(DescTools)
library(coin)
library(randtests)
library(kableExtra)

```

# EXERCISE 13.1

Ten students take two exams. Their scores are:

|             |     |     |     |     |     |     |     |     |     |     |
|-------------|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| **exam I**  | 65  | 70  | 55  | 85  | 75  | 80  | 90  | 60  | 95  | 70  |
| **exam II** | 60  | 68  | 50  | 82  | 72  | 78  | 88  | 58  | 92  | 65  |

## A. Compute Spearman’s rank correlation coefficient

```{r, echo = FALSE, message = FALSE, warning = FALSE}
exam <- data.frame("score" = c(65, 70, 55, 85, 75, 80, 90, 60, 95, 70,
                               60, 68, 50, 82, 72, 78, 88, 58, 92, 65),
                   "series" = rep(c("I","II"), each = 10)
                   )

# SPEARMAN'S RANK CORRELATION 
exam_cor <- cor.test(
                x = exam$score[exam$series == "I"],
                y = exam$score[exam$series == "II"],
                alternative = "two.sided",
                conf.level = 0.95,
                method = "spearman")

cat(
  paste0("Spearman's rank correlation","\n",
        round(exam_cor$estimate, 4))
)

```

## B. Test whether ρ is significantly different from zero at α = 0.05

```{r, echo = FALSE, message = FALSE, warning = FALSE}
cat(
  paste0("Spearman's rank correlation","\n",
         "p-value: ",round(exam_cor$p.value, 4),"\n",
         "Decision: ", ifelse(
           exam_cor$p.value < 0.05, 
           yes = "We reject null hypothesis. rho significant from zero",
           no = "We fail to reject the null hypothesis. rho is statistically    
           insignificant."))
)
```

## C. Interpret your results

In conclusion, there exists a significant monotonic relationship between the student performance of both examinations at a 5% level.

# EXERCISE 13.2

Using the data from Exercise 1, compute Kendall’s tau and test its significance. Compare with Spearman’s ρ

```{r, echo = FALSE, message = FALSE, warning = FALSE}
exam_cor2 <- cor.test(
                x = exam$score[exam$series == "I"],
                y = exam$score[exam$series == "II"],
                alternative = "two.sided",
                conf.level = 0.95,
                method = "kendall")

cat(
  paste0("Kendall's tau: ", round(exam_cor2$estimate, 4),"\n",
         "p-value: ", round(exam_cor2$p.value, 4),"\n",
         "Decision: ", ifelse(
           exam_cor2$p.value < 0.05, 
           yes = "We reject null hypothesis. tau significant from zero",
           no = "We fail to reject the null hypothesis. tau is statistically    
           insignificant."),"\n","\n","Comparison:","\n",
         "Both the tests draw similar estimate values and share the same conclusion of monotonic statistical significance between the exam scores for the two exams")
)

```

# EXERCISE 13.3

The following data contain ties

|       |     |     |     |     |     |     |     |     |     |     |
|-------|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| **X** | 2   | 3   | 3   | 4   | 5   | 5   | 5   | 6   | 7   | 8   |
| **Y** | 4   | 5   | 6   | 5   | 7   | 7   | 8   | 8   | 9   | 10  |

## A. Compute Spearman’s ρ

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ties <- data.frame("values" = c(2, 3, 3, 4, 5, 5, 5, 6, 7, 8,
                                4, 5, 6, 5, 7, 7, 8, 8, 9, 10),
                   "type" = rep(c("X","Y"), each = 5))

ties_cor <- cor.test(
  x = ties$values[ties$type == "X"],
  y = ties$values[ties$type == "Y"],
  alternative = "two.sided",
  conf.level = 0.95,
  method = "spearman")
  
cat(
  paste0("Spearman's rank correlation: ", round(ties_cor$estimate, 4))
)

```

## B. Compute Kendall’s τ

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ties_cor2 <- cor.test(
  x = ties$values[ties$type == "X"],
  y = ties$values[ties$type == "Y"],
  alternative = "two.sided",
  conf.level = 0.95,
  method = "kendall")
  
cat(
  paste0("Kendall's tau: ", round(ties_cor2$estimate, 4))
)
```

## C. Which measure do you think is more appropriate for these data?

The Kendall's tau. It is better at handling tied observations better than Spearman's rank correlation.

# EXERCISE 13.4

Generate a dataset of n = 10 with a perfect positive linear relationship (X = 1 : 10, Y = X). Then add an outlier (X = 11, Y = 1)

## A. Compute Pearson’s r, Spearman’s ρ, and Kendall’s τ for the original data

```{r, echo = FALSE, message = FALSE, warning = FALSE}
gen <- data.frame("values" = rep(1:10, times = 2),
                  "type" = rep(c("X", "Y"), each = 10)
                  )

# gen <- rbind(gen, c(11, "X"), c(1, "Y")); gen$values <- gen$values |> as.numeric() 

gen_cor <- cor.test(
  x = gen$values[gen$type == "X"],
  y = gen$values[gen$type == "Y"],
  alternative = "two.sided",
  conf.level = 0.95,
  method = "pearson")

gen_cor2 <- cor.test(
  x = gen$values[gen$type == "X"],
  y = gen$values[gen$type == "Y"],
  alternative = "two.sided",
  conf.level = 0.95,
  method = "spearman")

gen_cor3 <- cor.test(
  x = gen$values[gen$type == "X"],
  y = gen$values[gen$type == "Y"],
  alternative = "two.sided",
  conf.level = 0.95,
  method = "kendall")

cat(
  paste0("Pearson's correlation: ", round(gen_cor$estimate, 4),"\n",
         "Spearman's rank correlation: ", round(gen_cor2$estimate, 4), "\n",
         "Kendall's tau: ", round(gen_cor3$estimate, 4))
)

```

## B. Compute the same measures for the data with the outlier

```{r, echo = FALSE, message = FALSE, warning = FALSE}
gen2 <- rbind(gen, c(11, "X"), c(1, "Y")); gen2$values <- gen2$values |> as.numeric()

gen_cor4 <- cor.test(
  x = gen2$values[gen2$type == "X"],
  y = gen2$values[gen2$type == "Y"],
  alternative = "two.sided",
  conf.level = 0.95,
  method = "pearson")

gen_cor5 <- cor.test(
  x = gen2$values[gen2$type == "X"],
  y = gen2$values[gen2$type == "Y"],
  alternative = "two.sided",
  conf.level = 0.95,
  method = "spearman")

gen_cor6 <- cor.test(
  x = gen2$values[gen2$type == "X"],
  y = gen2$values[gen2$type == "Y"],
  alternative = "two.sided",
  conf.level = 0.95,
  method = "kendall")

cat(
  paste0("Pearson's correlation: ", round(gen_cor4$estimate, 4),"\n",
         "Spearman's rank correlation: ", round(gen_cor5$estimate, 4), "\n",
         "Kendall's tau: ", round(gen_cor6$estimate, 4))
)
```

## C. Compare the robustness of the three measures

Kendall's tau has the least deviation as compared to pearson and spearman correlation as a result of adding the extreme observation. This makes it the most robust out of the three.

# EXERCISE 13.5

For the data in Exercise 1, construct:

## A. A 95% confidence interval for Spearman's rho using Fisher's z-transformation

```{r, echo = FALSE, message = FALSE, warning = FALSE}
rho_z_value <- 0.5*log((1+exam_cor$estimate)/(1-exam_cor$estimate))
rho_sd <- 1/(sqrt(10-3))

z_lower <- rho_z_value - qnorm(0.975)*rho_sd
z_upper <- rho_z_value + qnorm(0.975)*rho_sd

# Back transforming
lower <- ((exp(2*z_lower)-1)/(exp(2*z_lower)+1)) |> round(4)

upper <- ((exp(2*z_upper)-1)/(exp(2*z_upper)+1)) |> round(4)

cat(paste0("[",lower,", ",upper,"]"))
```

## B. A 95% confidence interval for Kendall's tau using the large-scale approximation

```{r, echo = FALSE, message = FALSE, warning = FALSE}
n <- 10
kendall_z_variate <- qnorm(0.975)

tau_sd <- sqrt(
  (2*n*(n+5))/(9*n*(n-1))
  )

l2 <- max(-1, exam_cor2$estimate - kendall_z_variate*(tau_sd)) |> round(4)
u2 <- min(1, exam_cor2$estimate + kendall_z_variate*(tau_sd)) |> round(4)
paste0("[",l2,", ",u2,"]") |> cat()
```

## C. Compare the widths of the intervals

Spearman's rank correlation has a narrower interval than Kendall's tau.

# EXERCISE 13.6

A survey classifies 200 individuals by gender and political party preference:

|        |          |            |             |       |
|--------|----------|------------|-------------|-------|
|        | Democrat | Republican | Independent | Total |
| Male   | 45       | 55         | 20          | 120   |
| Female | 60       | 40         | 30          | 130   |
| Total  | 105      | 95         | 50          | 250   |

## A. Test whether gender and party preference are independent using the Chi-square test at 5% significance level

```{r, echo = FALSE, message = FALSE, warning = FALSE}
political <- matrix(c(45,55,20,60,40,30), nrow = 2, byrow = TRUE)

political_chi <- chisq.test(political)

cat(
  paste0("Chi-square test for independence", "\n",
         "p-value: ",round(political_chi$p.value, 4), "\n",
         "Decision: ", ifelse(political_chi$p.value < 0.05, yes = "Reject the null hypothesis", no = "Fail to reject the null hypothesis"))
)
```

## B. Check whether the chi-square approximation is valid

```{r, echo = FALSE, message = FALSE, warning = FALSE}
political_chi$expected |> matrix(nrow = 2,
                dimnames = list(c("Male", "Female"),
                                c("Democrat", "Republican", "Independent"))) 
```

No expected value is below 5. Thus, the chi-square approximation is valid.

## C. Interpret your results

At a 5% significance level, sample data from 250 participants show that the choice of political party is associated to gender.

# EXERCISE 13.7

Simulate data from the following models with $n = 20$:

## A

$Y=X + \epsilon, \epsilon \sim N(0,1), X\sim U(0,10)$

```{r, echo = FALSE, message = FALSE, warning = FALSE}
set.seed(314)
alpha <- 0.05
n_sim <- 1000

reject_pearson <- numeric(n_sim)
reject_spearman <- numeric(n_sim)
reject_kendall <- numeric(n_sim)


# Simulation A
for(i in 1:n_sim){
# Simulation
X <- runif(n = 20, min = 0, max = 10)
e <- rnorm(n = 20, mean = 0, sd = 1)
Y <- X + e

# Hypothesis testing
## Pearson r
p_pearson <- cor.test(
  x = X, y = Y, method = "pearson", alternative = "two.sided"
  )$p.value

reject_pearson[i] <- (p_pearson < alpha)

## Spearman rho 
p_spearman <- cor.test(
  x = X, y = Y, method = "spearman", alternative = "two.sided"
  )$p.value

reject_spearman[i] <- (p_spearman < alpha)

## Kendall tau
p_kendall <- cor.test(
  x = X, y = Y, method = "kendall", alternative = "two.sided"
  )$p.value

reject_kendall[i] <- (p_kendall < alpha)

}

# Empirical power
power_pearson <- mean(reject_pearson)
power_spearman <- mean(reject_spearman)
power_kendall <- mean(reject_kendall)

cat(
  paste0("Empirical power: Simulation A","\n",
         "Pearson: ",power_pearson*100, "%", "\n",
         "Spearman: ", power_spearman*100, "%", "\n",
         "Kendall: ", power_kendall*100, "%","\n","\n","\n"
  )
)
```

## B

$Y={X}^{2} + \epsilon, \epsilon \sim N(0,5), X \sim U(-3,3)$

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Simulation B
reject_pearson2 <- numeric(n_sim)
reject_spearman2 <- numeric(n_sim)
reject_kendall2 <- numeric(n_sim)

for(i in 1:n_sim){

X2 <- runif(n = 20, min = -3, max = 3)
e2 <- rnorm(n = 20, mean = 0, sd = 5)
Y2 <- X2^2 + e2

# Hypothesis testing
## Pearson r
p_pearson2 <- cor.test(
  x = X2, y = Y2, method = "pearson", alternative = "two.sided"
  )$p.value

reject_pearson2[i] <- (p_pearson2 < alpha)

## Spearman rho 
p_spearman2 <- cor.test(
  x = X2, y = Y2, method = "spearman", alternative = "two.sided"
  )$p.value

reject_spearman2[i] <- (p_spearman2 < alpha)

## Kendall tau
p_kendall2 <- cor.test(
  x = X2, y = Y2, method = "kendall", alternative = "two.sided"
  )$p.value

reject_kendall2[i] <- (p_kendall2 < alpha)

}

# Empirical power
power_pearson2 <- mean(reject_pearson2)
power_spearman2 <- mean(reject_spearman2)
power_kendall2 <- mean(reject_kendall2)

cat(
  paste0("Empirical power: Simulation B","\n",
         "Pearson: ",power_pearson2*100, "%", "\n",
         "Spearman: ", power_spearman2*100, "%", "\n",
         "Kendall: ", power_kendall2*100, "%","\n","\n","\n"
  )
)

```

## C

$Y={e}^{X} + \epsilon, \epsilon \sim N(0,1), X\sim U(0,2)$

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Simulation C
reject_pearson3 <- numeric(n_sim)
reject_spearman3 <- numeric(n_sim)
reject_kendall3 <- numeric(n_sim)

for(i in 1:n_sim){

X3 <- runif(n = 20, min = 0, max = 2)
e3 <- rnorm(n = 20, mean = 0, sd = 1)
Y3 <- exp(X3) + e3

# Hypothesis testing
## Pearson r
p_pearson3 <- cor.test(
  x = X3, y = Y3, method = "pearson", alternative = "two.sided"
  )$p.value

reject_pearson3[i] <- (p_pearson3 < alpha)

## Spearman rho 
p_spearman3 <- cor.test(
  x = X3, y = Y3, method = "spearman", alternative = "two.sided"
  )$p.value

reject_spearman3[i] <- (p_spearman3 < alpha)

## Kendall tau
p_kendall3 <- cor.test(
  x = X3, y = Y3, method = "kendall", alternative = "two.sided"
  )$p.value

reject_kendall3[i] <- (p_kendall3 < alpha)

}

# Empirical power
power_pearson3 <- mean(reject_pearson3)
power_spearman3 <- mean(reject_spearman3)
power_kendall3 <- mean(reject_kendall3)

cat(
  paste0("Empirical power: Simulation C","\n",
         "Pearson: ",power_pearson3*100, "%", "\n",
         "Spearman: ", power_spearman3*100, "%", "\n",
         "Kendall: ", power_kendall3*100, "%","\n","\n","\n"
  )
)

```

For each model, compute Pearson's r, Spearman's $\rho$ and Kendall's $\tau$. Repeat 1000 times and compare the empirical power to detect association. What do you conclude about the robustness of each measure to non-linearity

|          |                   |                  |                  |
|----------|-------------------|------------------|------------------|
|          | **simulation A**  | **simulation B** | **simulation C** |
| Pearson  | 100%              | 6.1%             | 100%             |
| Spearman | 100%              | 6.3%             | 99.7%            |
| Kendall  | 100%              | 7%               | 99.9%            |

**Pearson's coefficient** is robust to variables that exhibit linear (simulation A) and exponential (simulation C) associations. It performs poorly when trying to capture associations between variables having a quadratic trend. (simulation B)

**Spearman's rank coefficient** is more robust when it makes an attempt to capture linear associations as compared to exponential ones. It registers a poor but slightly better empirical power than Pearson's coefficient when capturing quadratic associations.

The **Kendall's tau** mirrors the Spearman rank coeffient in terms of the robustness property in capturing linear and exponential relationships. It performs better when attempting to capture quadratic relationships ahead of Pearson and Spearman coefficients.
