---
title: "Non-parametric Methods - Problems"
author: "Gift Pamba"
date: "April-May 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: darkly
  word_document:
    toc: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(DescTools)
library(kableExtra)

```

# Comparing two related samples

## ONE

A teacher wished to determine if providing a bilingual dictionary to students with limited English proficiency improves math test scores. A small class of students **(n = 10)** was selected. Students were given two math tests. Each test covered the same type of math content; however, students were provided a bilingual dictionary on the second test. The data in `Table 3.10` represent the students’ performance on each math test.

| student | Math test without a billingual dictionary | Math test with a billingual dictionary |
|------------------|----------------------------|--------------------------|
| 1 | 30 | 39 |
| 2 | 56 | 46 |
| 3 | 48 | 37 |
| 4 | 47 | 44 |
| 5 | 43 | 32 |
| 6 | 45 | 39 |
| 7 | 36 | 41 |
| 8 | 44 | 40 |
| 9 | 44 | 38 |
| 10 | 40 | 46 |

: Table 3.10

Use a **one-tailed Wilcoxon signed rank** test and a **one-tailed sign test** to determine which testing condition resulted in higher scores. Use $\alpha = 0.05$. Report your findings

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tab1 <- data.frame(
  "student"=seq(1,10,1),
  "before"=c(30,56,48,47,43,45,36,44,44,40),
  "after"=c(39,46,37,44,32,39,41,40,38,46)
)
```

### a

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dictionary <- wilcox.test(x = tab1$before, y = tab1$after,
                          alternative = "greater",
                          paired = TRUE)
```

### b

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dictionary2 <- SignTest(x = tab1$before, y = tab1$after,
                        alternative = "greater")
```

## TWO

A research study was done to investigate the influence of being alone at night on the human male heart rate.Two mean were sent into a wooded area, one at a time, at night, for 20 minutes. They had a heart monitor to record their pulse rate. The second night, the same men were sent into a similar wooded area accompanied by a companion. Thei pulse rate was recorded again. The researcher wanted to see if having a companion would change their pulse rate. The median pulse rates are recorded in `Table 3.11`. Use a **two-tailed Wilcoxon signed-rank test** and a **two-tailed signed test** to determine which condition produced a higher pulse rate. Use $\alpha = 0.05$. Report your findings

| participant | median rate alone | median rate with companion |
|-------------|-------------------|----------------------------|
| A           | 88                | 72                         |
| B           | 77                | 74                         |
| C           | 91                | 80                         |
| D           | 70                | 77                         |
| E           | 80                | 71                         |
| F           | 85                | 83                         |
| G           | 90                | 80                         |
| H           | 82                | 91                         |
| I           | 93                | 86                         |
| J           | 75                | 69                         |

: Table 3.11

```{r, echo=FALSE, message=FALSE, warning=FALSE}
pulse <- data.frame(
  "participant" = LETTERS[1:10],
  "without" = c(88,77,91,70,80,85,90,82,93,75),
  "with" = c(72,74,80,77,71,83,80,91,86,69))
```

### a

```{r, echo=FALSE, message=FALSE, warning=FALSE}
hr <- wilcox.test(x = pulse$without, y = pulse$with,
                  alternative = "two.sided", 
                  paired = TRUE)
```

### b

```{r, echo=FALSE, message=FALSE, warning=FALSE}
hr2 <- SignTest(x = pulse$without, y = pulse$with,
                  alternative = "two.sided")
```

## THREE

A researcher conducts a pilot study to compare two treatments to help obese female teenagers lose weight. She tests each individual in two different treatment conditions. The data in `Table 3.12` provides the number of pounds that each participant lost.

|                 |                 |                 |
|-----------------|-----------------|-----------------|
| **participant** | **treatment 1** | **treatment 2** |
| 1               | 10              | 18              |
| 2               | 20              | 12              |
| 3               | 15              | 16              |
| 4               | 9               | 7               |
| 5               | 18              | 21              |
| 6               | 11              | 17              |
| 7               | 6               | 13              |
| 8               | 12              | 14              |

: Table 3.12

Use a **two-tailed Wilcoxon signed-rank test** and a **two-tailed sign test** to determine which treatment resulted in greater weight loss. Use $\alpha = 0.05$. Report your findings

```{r, echo = FALSE, message = FALSE, warning = FALSE}
wl <- data.frame(
  "participant" = 1:8,
  "t1" = c(10,20,15,9,18,11,6,12),
  "t2" = c(18,12,16,7,21,17,13,14))
```

### a

```{r, echo = FALSE, message = FALSE, warning = FALSE}
wl_wsr <- wilcox.test(x = wl$t1, y = wl$t2,
                      alternative = "two.sided", paired = TRUE)
```

### b

```{r, echo = FALSE, message = FALSE, warning = FALSE}
wl_sign <- SignTest(x = wl$t1, y = wl$t2,
                    alternative = "two.sided")
```

## FOUR

Twenty participants in an exercise program were measured on the number of sit-ups they could do before physical exercise (first count) and the number they could do after they had done at least 45 minutes of other physical exercise.(second count) `Table 3.13` shows the results for 20 participants obbtained during two separate physical exercise sessions. Determine the **Effect Size** for a calculated z-score

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ex <- data.frame(
"participant" = seq(1,20,1),
"first" = c(18,19,20,29,15,22,21,30,22,11,20,21,21,20,18,27,24,13,10,10),
"second" = c(28,18,28,20,30,25,28,18,27,30,24,27,10,40,20,14,29,30,24,36)
)

# Sample size
n_ex <- nrow(ex)

# Wilcoxon signed-rank hypothesis test for paired samples
ex_wsr <- wilcox.test(x = ex$first, y = ex$second,
                      alternative = "two.sided",
                      paired = TRUE)

# Test statistic
ex_wsr_t <- ex_wsr$statistic

# Normal approximation
## E[T]
ex_wsr_tmean <- (n_ex*n_ex+1)/4

## sd[T]
ex_wsr_tsd <- sqrt((n_ex*(n_ex+1)*(2*n_ex+1))/24)

## z-variate
z_ex <- ((ex_wsr_t-ex_wsr_tmean)/ex_wsr_tsd)

## Effect size
es_ex <- abs(z_ex)/sqrt(n_ex)

cat(paste0("Effect size: ", round(es_ex, 4)))
```

## FIVE

A school is trying to get more students to participate in activities that will make learning more desirable. `Table 3.14` shows the number of activities that each of the 10 students in one class participated in last year before a new activity program was implemented and this year after it was implemented. Construct a 95% confidence interval based on the **Wilcoxon signed-rank test** to determine whether the new activity program had a significant positive effect on the student participation

```{r, echo=FALSE, message=FALSE, warning=FALSE}
activity <- data.frame(
  "participant" = seq(1,10,1),
  "lastyr" = c(18,22,10,25,16,14,21,13,28,12),
  "thisyr" = c(20,28,18,23,20,21,17,18,22,21)
)

# Wilcoxon signed-rank hypothesis test
act_wsr <- wilcox.test(x = activity$lastyr, y = activity$thisyr,
                       alternative = "two.sided", paired = TRUE,
                       conf.int = TRUE)

cat(
  paste0("[",round(act_wsr$conf.int[1],4),",",round(act_wsr$conf.int[2],4),"]")
)
```

# Comparing two unrelated samples

## ONE

The data in `Table 4.8` were obtained from a reading-level test for 1st grade children. Compare the performance gains of the two different methods for teaching reading.

| Method     | Gain score | Method      | Gain score |
|------------|------------|-------------|------------|
| One on one | 16         | Small group | 11         |
| One on one | 13         | Small group | 2          |
| One on one | 16         | Small group | 10         |
| One on one | 16         | Small group | 4          |
| One on one | 13         | Small group | 9          |
| One on one | 9          | Small group | 8          |
| One on one | 12         | Small group | 5          |
| One on one | 12         | Small group | 6          |
| One on one | 20         | Small group | 4          |
| One on one | 17         | Small group | 16         |

: Table 4.8

Use **two-tailed Mann-Whitney U** and **Kolmogorov-Smirnov two-sample** tests to determine which method was better for teaching reading. Set $\alpha = 0.05$. Report your findings

### a

### b

## TWO

A research study was conducted to see if an active involvement in a hobby has a positive effect on the health of a person who retires after age 65. The data in `Table 4.9` describes the health (number of doctor visits in 1 year) for participants who are involved in a hobby almost daily and those who are not.

| No hobby group | Hobby group |
|----------------|-------------|
| 12             | 9           |
| 15             | 5           |
| 8              | 10          |
| 11             | 3           |
| 9              | 4           |
| 17             | 2           |

: Table 4.9

Use **one-tailed Mann-Whitney U** and **Kolmogorov-Smirnov two-sample** tests to determine whether the hobby tends to reduce the need for doctor visits. Set $\alpha = 0.05$. Report your findings

### a

### b

| Method 1 | Method 2 |
|----------|----------|
| 53       | 91       |
| 41       | 18       |
| 17       | 14       |
| 45       | 21       |
| 44       | 23       |
| 12       | 99       |
| 49       | 16       |
| 50       | 10       |

: Table 4.10

## THREE

`Table 4.10` shows assessment scores of two different classes who are being taught computer skills using two different methods

Use **two-tailed Mann-Whitney U** and **Kolmogorov-Smirnov two-sample** tests to determine which method was better for teaching compute skills. Set $\alpha = 0.05$. Report your findings

### a

### b

## FOUR

Two methods of teaching were compared. Method 1 uses the computer to interact with the student and diagnose and remediate the student based on misconceptions. Method 2 was taught using workbooks in classroom groups. `Table 4.11` shows the data obtained on an assessment after 6 weeks of instruction. Calculate the ES using the z-score from the analysis.

|          |          |
|----------|----------|
| Method 1 | Method 2 |
| 27       | 9        |
| 38       | 42       |
| 15       | 21       |
| 85       | 83       |
| 36       | 110      |
| 95       | 19       |
| 93       | 29       |
| 57       | 40       |
| 63       | 30       |
| 81       | 23       |
| 65       | 18       |
| 77       | 32       |
| 59       | 101      |
| 89       | 7        |
| 41       | 50       |
| 26       | 37       |
| 102      | 22       |
| 55       | 71       |
| 46       | 16       |
| 82       | 45       |
| 24       | 35       |
| 87       | 28       |
| 66       | 91       |
| 12       | 86       |
| 90       | 20       |

: Table 4.11

## FIVE

Two methods were used to provide instruction in Scienc efor 7th Grade. Method 1 included a laboratory each week and method 2 has only classroom work with lecture and worksheets. `Table 4.12` shows end-of-term test performance for the two methods. Construct a 95% median confidence interval based on the difference between two independent samples to compare the two methods.

| Method 1 | Method 2 |
|----------|----------|
| 15       | 8        |
| 23       | 15       |
| 9        | 10       |
| 12       | 13       |
| 18       | 17       |
| 22       | 5        |
| 17       | 18       |
| 20       | 7        |

: Table 4.12

# Comparing more than two related samples

## ONE

A graduate student performed a pilot study for his dissertation. He wanted to examine the effects of animal companionship on elderly males. He selected 10 male participants from a nursing home. Then he used an ABAB research design, where A represented a week with the absence of a cat and B represented a week with the presence of a cat. At the end of each week, he administered a 20-point survey to measure the quality of life satisfaction. The survey results are presented in `Table 5.9`

| participant | week 1 | week 2 | week 3 | week 4 |
|-------------|--------|--------|--------|--------|
| 1           | 7      | 6      | 8      | 9      |
| 2           | 9      | 8      | 10     | 1      |
| 3           | 15     | 18     | 16     | 17     |
| 4           | 7      | 6      | 8      | 9      |
| 5           | 7      | 8      | 10     | 11     |
| 6           | 10     | 14     | 13     | 11     |
| 7           | 12     | 19     | 11     | 13     |
| 8           | 7      | 4      | 2      | 5      |
| 9           | 8      | 7      | 9      | 5      |
| 10          | 12     | 16     | 14     | 15     |

: Table 5.9

Use a **Friedman test** to determine if one or more of the groups are significantly different. Since this is a pilot study, use $\alpha = 0.10$. If a significant difference exists, use **Wilcoxon signed-rank tests** to identify which groups are significantly different. Use the **Bonferroni procedure** to limit the type I error rate. Report your findings

## TWO

A physical education teacher conducted an action research project to examine a strength and conditioning program. Using 12 male participants, she measures the number of curl ups they could do in 1 minute. She measured their performance before the program. Then, she measured their performance at 1 month intervals. `Table 5.10` presents the performance results.

| participant | baseline  | month 1 | month 2 |
|-------------|-----------|---------|---------|
| 1           | 66        | 67      | 69      |
| 2           | 49        | 50      | 56      |
| 3           | 51        | 52      | 49      |
| 4           | 65        | 65      | 69      |
| 5           | 42        | 43      | 46      |
| 6           | 38        | 39      | 40      |
| 7           | 33        | 31      | 39      |
| 8           | 41        | 41      | 44      |
| 9           | 46        | 47      | 48      |
| 10          | 45        | 46      | 46      |
| 11          | 36        | 33      | 34      |
| 12          | 51        | 55      | 67      |

: Table 5.10: Number of curl ups in a minute


Use a **Friedman test** with $\alpha = 0.05$ to determine if one or more of the groups are significantly different. The teacher is expecting performance gains, so if a significant difference exists, use **one-tailed Wilcoxon signed-rank tests** to identify which groups are significantly different. Use the **Bonferroni** procedure to limit the type I error rate. Report your findings.  