---
title: "Non-parametric Methods - Problems"
author: "Gift Pamba"
date: "May-June 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: darkly
    code_folding: show
  word_document:
    toc: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(DescTools)
library(kableExtra)

```

# 1. Comparing two related samples

## ONE

A teacher wished to determine if providing a bilingual dictionary to students with limited English proficiency improves math test scores. A small class of students **(n = 10)** was selected. Students were given two math tests. Each test covered the same type of math content; however, students were provided a bilingual dictionary on the second test. The data in `Table 3.10` represent the students’ performance on each math test.

| student | Math test without a billingual dictionary | Math test with a billingual dictionary |
|-------------------|---------------------------|--------------------------|
| 1 | 30 | 39 |
| 2 | 56 | 46 |
| 3 | 48 | 37 |
| 4 | 47 | 44 |
| 5 | 43 | 32 |
| 6 | 45 | 39 |
| 7 | 36 | 41 |
| 8 | 44 | 40 |
| 9 | 44 | 38 |
| 10 | 40 | 46 |

: Table 3.10

Use a **one-tailed Wilcoxon signed rank** test and a **one-tailed sign test** to determine which testing condition resulted in higher scores. Use $\alpha = 0.05$. Report your findings

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tab1 <- data.frame(
  "student"=seq(1,10,1),
  "before"=c(30,56,48,47,43,45,36,44,44,40),
  "after"=c(39,46,37,44,32,39,41,40,38,46)
)
```

### a

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dictionary <- wilcox.test(x = tab1$before, y = tab1$after,
                          alternative = "greater",
                          paired = TRUE)
```

### b

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dictionary2 <- SignTest(x = tab1$before, y = tab1$after,
                        alternative = "greater")
```

## TWO

A research study was done to investigate the influence of being alone at night on the human male heart rate.Two mean were sent into a wooded area, one at a time, at night, for 20 minutes. They had a heart monitor to record their pulse rate. The second night, the same men were sent into a similar wooded area accompanied by a companion. Thei pulse rate was recorded again. The researcher wanted to see if having a companion would change their pulse rate. The median pulse rates are recorded in `Table 3.11`. Use a **two-tailed Wilcoxon signed-rank test** and a **two-tailed signed test** to determine which condition produced a higher pulse rate. Use $\alpha = 0.05$. Report your findings

| participant | median rate alone | median rate with companion |
|-------------|-------------------|----------------------------|
| A           | 88                | 72                         |
| B           | 77                | 74                         |
| C           | 91                | 80                         |
| D           | 70                | 77                         |
| E           | 80                | 71                         |
| F           | 85                | 83                         |
| G           | 90                | 80                         |
| H           | 82                | 91                         |
| I           | 93                | 86                         |
| J           | 75                | 69                         |

: Table 3.11

```{r, echo=FALSE, message=FALSE, warning=FALSE}
pulse <- data.frame(
  "participant" = LETTERS[1:10],
  "without" = c(88,77,91,70,80,85,90,82,93,75),
  "with" = c(72,74,80,77,71,83,80,91,86,69))
```

### a

```{r, echo=FALSE, message=FALSE, warning=FALSE}
hr <- wilcox.test(x = pulse$without, y = pulse$with,
                  alternative = "two.sided", 
                  paired = TRUE)
```

### b

```{r, echo=FALSE, message=FALSE, warning=FALSE}
hr2 <- SignTest(x = pulse$without, y = pulse$with,
                  alternative = "two.sided")
```

## THREE

A researcher conducts a pilot study to compare two treatments to help obese female teenagers lose weight. She tests each individual in two different treatment conditions. The data in `Table 3.12` provides the number of pounds that each participant lost.

|                 |                 |                 |
|-----------------|-----------------|-----------------|
| **participant** | **treatment 1** | **treatment 2** |
| 1               | 10              | 18              |
| 2               | 20              | 12              |
| 3               | 15              | 16              |
| 4               | 9               | 7               |
| 5               | 18              | 21              |
| 6               | 11              | 17              |
| 7               | 6               | 13              |
| 8               | 12              | 14              |

: Table 3.12

Use a **two-tailed Wilcoxon signed-rank test** and a **two-tailed sign test** to determine which treatment resulted in greater weight loss. Use $\alpha = 0.05$. Report your findings

```{r, echo = FALSE, message = FALSE, warning = FALSE}
wl <- data.frame(
  "participant" = 1:8,
  "t1" = c(10,20,15,9,18,11,6,12),
  "t2" = c(18,12,16,7,21,17,13,14))
```

### a

```{r, echo = FALSE, message = FALSE, warning = FALSE}
wl_wsr <- wilcox.test(x = wl$t1, y = wl$t2,
                      alternative = "two.sided", paired = TRUE)
```

### b

```{r, echo = FALSE, message = FALSE, warning = FALSE}
wl_sign <- SignTest(x = wl$t1, y = wl$t2,
                    alternative = "two.sided")
```

## FOUR

Twenty participants in an exercise program were measured on the number of sit-ups they could do before physical exercise (first count) and the number they could do after they had done at least 45 minutes of other physical exercise.(second count) `Table 3.13` shows the results for 20 participants obbtained during two separate physical exercise sessions. Determine the **Effect Size** for a calculated z-score

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ex <- data.frame(
"participant" = seq(1,20,1),
"first" = c(18,19,20,29,15,22,21,30,22,11,20,21,21,20,18,27,24,13,10,10),
"second" = c(28,18,28,20,30,25,28,18,27,30,24,27,10,40,20,14,29,30,24,36)
)

# Sample size
n_ex <- nrow(ex)

# Wilcoxon signed-rank hypothesis test for paired samples
ex_wsr <- wilcox.test(x = ex$first, y = ex$second,
                      alternative = "two.sided",
                      paired = TRUE)

# Test statistic
ex_wsr_t <- ex_wsr$statistic

# Normal approximation
## E[T]
ex_wsr_tmean <- ((n_ex)*(n_ex+1))/4

## sd[T]
ex_wsr_tsd <- sqrt((n_ex*(n_ex+1)*(2*n_ex+1))/24)

## z-variate
z_ex <- ((ex_wsr_t-ex_wsr_tmean)/ex_wsr_tsd)

## Effect size
es_ex <- abs(z_ex)/sqrt(n_ex)

cat(paste0("Effect size: ", round(es_ex, 4)))
```

## FIVE

A school is trying to get more students to participate in activities that will make learning more desirable. `Table 3.14` shows the number of activities that each of the 10 students in one class participated in last year before a new activity program was implemented and this year after it was implemented. Construct a 95% confidence interval based on the **Wilcoxon signed-rank test** to determine whether the new activity program had a significant positive effect on the student participation

```{r, echo=FALSE, message=FALSE, warning=FALSE}
activity <- data.frame(
  "participant" = seq(1,10,1),
  "lastyr" = c(18,22,10,25,16,14,21,13,28,12),
  "thisyr" = c(20,28,18,23,20,21,17,18,22,21)
)

# Wilcoxon signed-rank hypothesis test
act_wsr <- wilcox.test(x = activity$lastyr, y = activity$thisyr,
                       alternative = "two.sided", paired = TRUE,
                       conf.int = TRUE)

cat(
  paste0("[",round(act_wsr$conf.int[1],4),",",round(act_wsr$conf.int[2],4),"]")
)
```

## SIX

*(ANSM 4.3)*

Samples of cream from each of ten dairies (A to J) are each divided into two portions. One portion from each is sent to laboratory I, the other to Laboratory II, for bacterium counts. The counts (thousands bacteria ${ml}^{-1}$) are:

| Dairy | Lab I | Lab II |
|-------|-------|--------|
| A     | 11.7  | 10.9   |
| B     | 12.1  | 11.9   |
| C     | 13.3  | 13.4   |
| D     | 15.1  | 15.4   |
| E     | 15.9  | 14.8   |
| F     | 15.3  | 14.8   |
| G     | 11.9  | 12.3   |
| H     | 16.2  | 15.0   |
| I     | 15.1  | 14.2   |
| J     | 13.6  | 13.1   |

```{r, echo = FALSE, message = FALSE, warning = FALSE}
bacteria <- data.frame(
  "dairy" = LETTERS[1:10],
  "l1" = c(11.7,12.1,13.3,15.1,15.9,15.3,11.9,16.2,15.1,13.6),
  "l2" = c(10.9,11.9,13.4,15.4,14.8,14.8,12.3,15,14.2,13.1)
)
```

Use the **Wilcoxon signed-rank test** to assess the evidence for any consistent difference between laboratories for subsamples from the same dairy. Obtain also 95 and 99 percent confidence intervals for the mean and compare these with the intervals using the optimal method when normality is assumed

```{r, echo = FALSE, message = FALSE, warning = FALSE}
bacteria_wsr <- wilcox.test(x = bacteria$l1, y = bacteria$l2,
                          paired = TRUE, exact = FALSE, conf.int = TRUE,
                          alternative = "two.sided", conf.level = c(0.95))
```

## SEVEN

*(ANSM 4.6)*

For each of nine matched pairs of students, one student is allocated to a series of lectures and the other to appropriate computer assisted learning (CAL) material. At the end of the course, the students are given the same examination paper. The marks achieved (out of 100) are:

| pair | CAL | lectures |
|------|-----|----------|
| 1    | 50  | 25       |
| 2    | 56  | 58       |
| 3    | 51  | 65       |
| 4    | 46  | 38       |
| 5    | 88  | 91       |
| 6    | 79  | 32       |
| 7    | 81  | 31       |
| 8    | 95  | 13       |
| 9    | 73  | 49       |

```{r, echo = FALSE, message = FALSE, warning = FALSE}
lvc <- data.frame(
  "pair" = 1:9,
  "CAL" = c(50,56,51,46,88,79,81,95,73),
  "lectures" = c(25,58,65,38,91,32,31,13,49)
)
```

Analyze these results by what you consider the most appropriate parametric or nonparametric methods to determine whether or not they provide acceptable evidence that CAL materials lead to better examinations results

## EIGHT

*(ANSM 4.11)*

To produce high-quality steel one of two hardening agents A, B may be added to molten metal. Hardness of steel varies from batch to batch, so to test the two agents, batches are sub-divided into two portions, for each batch agent A being added to one portion, agent B to the other. To compare hardness, sharpened specimens for each pair are used to make scratches on each other; that making the deeper scratch on the other is the harder specimen. For a subsample of 10 pairs, an expensive test is used to produce a hardness index. The higher the value of the index, the harder the steel. The indices recorded were:

| batch no. | additive A | additive B |
|-----------|------------|------------|
| 1         | 22         | 27         |
| 2         | 26         | 25         |
| 3         | 29         | 31         |
| 4         | 22         | 27         |
| 5         | 31         | 29         |
| 6         | 34         | 41         |
| 7         | 31         | 32         |
| 8         | 20         | 27         |
| 9         | 33         | 32         |
| 10        | 34         | 34         |

```{r,echo = FALSE, message = FALSE, warning = FALSE}
molten <- data.frame(
  "batch"=1:10,
  "A" = c(22,26,29,22,31,34,31,20,33,34),
  "B" = c(27,25,31,27,29,41,32,27,32,34)
)
```

Use an appropriate test to determine whether these data support the initial claim of equal hardness.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
molten_wsr <- wilcox.test(x = molten$A, y = molten$B,
                          alternative = "two.sided",exact = FALSE, paired = TRUE)
```

## NINE

*(ANSM 4.12)*

On the day of the third round of the Open Golf Championship in 1987 before play started a television commentator said that conditions were such that the average scores of players were likely to be at least three higher than those for the second round. For a random sample of 10 of the 77 players participating in both rounds the scores were:

| player | round 2 | round 3 |
|--------|---------|---------|
| A      | 73      | 72      |
| B      | 73      | 79      |
| C      | 74      | 79      |
| D      | 66      | 77      |
| E      | 71      | 83      |
| F      | 73      | 78      |
| G      | 68      | 70      |
| H      | 72      | 78      |
| I      | 73      | 78      |
| J      | 72      | 77      |

```{r, echo = FALSE, message = FALSE, warning = FALSE}
golf <- data.frame(
  "player" = LETTERS[1:10],
  "r2" = c(73,73,74,66,71,73,68,72,73,72),
  "r3" = c(72,79,79,77,83,78,70,78,78,77)
)

```

Do these data support the commentator’s claim? Consider carefully whether a one- or two-tail test is appropriate

## TEN

*(ANSM 4.13)*

Pearson and Sprent (1968) gave data for hearing loss (in decibels below prescribed norms) at various frequencies. The data below show these losses for $10$ individuals aged between 46 and $54$ at frequencies of $0.125$ and $0.25\ kcs^-1$. A negative loss indicates hearing above the norm. Is there an indication of a different loss at the two frequencies

| subject | $0.125\ kcs^-1$ | $0.25\ kcs^-1$ |
|---------|-----------------|----------------|
| A       | 2.5             | 2.5            |
| B       | -7.5            | -5             |
| C       | 11.25           | 6.35           |
| D       | 7.5             | 6.25           |
| E       | 10              | 7.5            |
| F       | 5               | 3.75           |
| G       | 7.5             | 1.25           |
| H       | 2.5             | 0              |
| I       | 5               | 2.5            |
| J       | 8.75            | 5              |

## ELEVEN

*(ANSM 4.15)*

Scott, Smith and Jones (1977) give a table of estimates of the percentages of UK electors predicted to vote Conservative by two opinion polling organizations, A and B, in each month in the years 1965–70. For a random sample of 15 months during that period the paired percentages were:

| A    | B    |
|------|------|
| 43.5 | 45.5 |
| 51.2 | 44.5 |
| 46.8 | 45   |
| 55.5 | 54.5 |
| 45.5 | 49.5 |
| 42   | 43.5 |
| 36   | 41   |
| 49.8 | 53   |
| 42.5 | 48   |
| 50.8 | 52.5 |
| 36.6 | 41   |
| 47.6 | 47.5 |
| 41.9 | 42.5 |
| 48.4 | 45   |
| 53.5 | 52.5 |

Do these results indicate a significant tendency for one of the organizations to return higher percentages than the other? Obtain an appropriate 95 percent confidence interval for any mean or median difference between predictions during the period covered

------------------------------------------------------------------------

# 2. Comparing two unrelated samples

## ONE

The data in `Table 4.8` were obtained from a reading-level test for 1st grade children. Compare the performance gains of the two different methods for teaching reading.

| Method     | Gain score | Method      | Gain score |
|------------|------------|-------------|------------|
| One on one | 16         | Small group | 11         |
| One on one | 13         | Small group | 2          |
| One on one | 16         | Small group | 10         |
| One on one | 16         | Small group | 4          |
| One on one | 13         | Small group | 9          |
| One on one | 9          | Small group | 8          |
| One on one | 12         | Small group | 5          |
| One on one | 12         | Small group | 6          |
| One on one | 20         | Small group | 4          |
| One on one | 17         | Small group | 16         |

: Table 4.8

```{r, echo = FALSE, message = FALSE, warning = FALSE}
firstgrade <- data.frame(
  "method" = c(rep("ono", 10), rep("sg", 10)),
  "score" = c(16,13,16,16,13,9,12,12,20,17,
              11,2,10,4,9,8,5,6,4,16)
)
```

Use **two-tailed Mann-Whitney U** and **Kolmogorov-Smirnov two-sample** tests to determine which method was better for teaching reading. Set $\alpha = 0.05$. Report your findings

### a

```{r, echo = FALSE, message = FALSE, warning = FALSE}
fg_mw <- wilcox.test(x = firstgrade$score[firstgrade$method=="ono"],
                     y = firstgrade$score[firstgrade$method=="sg"],
                     alternative = "two.sided", paired = FALSE)
```

### b

```{r, echo = FALSE, message = FALSE, warning = FALSE}
fg_ks <- ks.test(x = firstgrade$score[firstgrade$method=="ono"],
                 y = firstgrade$score[firstgrade$method=="sg"],
                     alternative = "two.sided")
```

## TWO

A research study was conducted to see if an active involvement in a hobby has a positive effect on the health of a person who retires after age 65. The data in `Table 4.9` describes the health (number of doctor visits in 1 year) for participants who are involved in a hobby almost daily and those who are not.

| No hobby group | Hobby group |
|----------------|-------------|
| 12             | 9           |
| 15             | 5           |
| 8              | 10          |
| 11             | 3           |
| 9              | 4           |
| 17             | 2           |

: Table 4.9

```{r, echo = FALSE, message = FALSE, warning = FALSE}
hobby <- data.frame(
  "group" = c(rep("nohobby",6), rep("hobby",6)),
  "visits" = c(12,15,8,11,9,17,
               9,5,10,3,4,2)
)
```

Use **one-tailed Mann-Whitney U** and **Kolmogorov-Smirnov two-sample** tests to determine whether the hobby tends to reduce the need for doctor visits. Set $\alpha = 0.05$. Report your findings

### a

```{r, echo = FALSE, message = FALSE, warning = FALSE}
hobby_mw <- wilcox.test(x = hobby$visits[hobby$group=="nohobby"],
                        y = hobby$visits[hobby$group=="hobby"],
                        alternative = "greater", paired = FALSE)
```

### b

```{r, echo = FALSE, message = FALSE, warning = FALSE}
hobby_ks <- ks.test(x = hobby$visits[hobby$group=="nohobby"],
                    y = hobby$visits[hobby$group=="hobby"],
                    alternative = "two.sided")
```

## THREE

| Method 1 | Method 2 |
|----------|----------|
| 53       | 91       |
| 41       | 18       |
| 17       | 14       |
| 45       | 21       |
| 44       | 23       |
| 12       | 99       |
| 49       | 16       |
| 50       | 10       |

: Table 4.10

`Table 4.10` shows assessment scores of two different classes who are being taught computer skills using two different methods

```{r, echo = FALSE, message = FALSE, warning = FALSE}
compskills <- data.frame(
  "method" = c(rep("method1", 8), rep("method2", 8)),
  "score" = c(53,41,17,45,44,12,49,50,
               91,18,14,21,23,99,16,10)
  )
```

Use **two-tailed Mann-Whitney U** and **Kolmogorov-Smirnov two-sample** tests to determine which method was better for teaching compute skills. Set $\alpha = 0.05$. Report your findings

### a

```{r, echo = FALSE, message = FALSE, warning = FALSE}
comp_mw <- wilcox.test(x = compskills$score[compskills$method == "method1"],
                       y = compskills$score[compskills$method == "method2"],
                       alternative = "two.sided", paired = FALSE)
```

### b

```{r, echo = FALSE, message = FALSE, warning = FALSE}
comp_ks <- ks.test(x = compskills$score[compskills$method == "method1"],
                   y = compskills$score[compskills$method == "method2"],
                   alternative = "two.sided")
```

## FOUR

Two methods of teaching were compared. Method 1 uses the computer to interact with the student and diagnose and remediate the student based on misconceptions. Method 2 was taught using workbooks in classroom groups. `Table 4.11` shows the data obtained on an assessment after 6 weeks of instruction. Calculate the ES using the z-score from the analysis.

|          |          |
|----------|----------|
| Method 1 | Method 2 |
| 27       | 9        |
| 38       | 42       |
| 15       | 21       |
| 85       | 83       |
| 36       | 110      |
| 95       | 19       |
| 93       | 29       |
| 57       | 40       |
| 63       | 30       |
| 81       | 23       |
| 65       | 18       |
| 77       | 32       |
| 59       | 101      |
| 89       | 7        |
| 41       | 50       |
| 26       | 37       |
| 102      | 22       |
| 55       | 71       |
| 46       | 16       |
| 82       | 45       |
| 24       | 35       |
| 87       | 28       |
| 66       | 91       |
| 12       | 86       |
| 90       | 20       |

: Table 4.11

```{r, echo = FALSE, message = FALSE, warning = FALSE}
comp2 <- data.frame(
  "method" = c(rep("method1",25), rep("method2", 25)),
"score" = c(27,38,15,85,36,95,93,57,63,81,65,77,59,89,41,26,102,55,46,82,24,87,66,12,90,
9,42,21,83,110,19,29,40,30,23,18,32,101,7,50,37,22,71,16,45,35,28,91,86,20)
)

# Sample size
n_comp2 <- 25

# Wilcoxon rank sum hypothesis test for paired samples
comp2_wsr <- wilcox.test(x = comp2$score[comp2$method == "method1"], 
                         y = comp2$score[comp2$method == "method2"],
                      alternative = "two.sided", paired = FALSE)

# Test statistic
comp2_wsr_t <- comp2_wsr$statistic

# Normal approximation
## E[T]
comp2_wsr_tmean <- ((n_comp2)*(n_comp2+1))/4

## sd[T]
comp2_wsr_tsd <- sqrt((n_comp2*(n_comp2+1)*(2*n_comp2+1))/24)

## z-variate
z_comp2 <- ((comp2_wsr_t-comp2_wsr_tmean)/comp2_wsr_tsd)

## Effect size
es_comp2 <- abs(z_comp2)/sqrt(n_comp2)

cat(paste0("Effect size: ", round(es_comp2, 4)))
```

## FIVE

Two methods were used to provide instruction in Science for 7th Grade. Method 1 included a laboratory each week and method 2 has only classroom work with lecture and worksheets. `Table 4.12` shows end-of-term test performance for the two methods. Construct a 95% median confidence interval based on the difference between two independent samples to compare the two methods.

| Method 1 | Method 2 |
|----------|----------|
| 15       | 8        |
| 23       | 15       |
| 9        | 10       |
| 12       | 13       |
| 18       | 17       |
| 22       | 5        |
| 17       | 18       |
| 20       | 7        |

: Table 4.12

```{r, echo = FALSE, message = FALSE, warning = FALSE}
lab <- data.frame(
  "method" = c(rep("1", 8), rep("2", 8)),
  "score" = c(15,23,9,12,18,22,17,20,
              8,15,10,13,17,5,18,7)
)
```

## SIX

*(ANSM 5.5)*

An alloy is composed of zinc, copper and tin. It may be made at one of two temperatures H (higher) or L (lower). We wish to know if one temperature produces a harder alloy. A sample is taken from each of 9 batches at L and 7 at H. To arrange them in ascending order of hardness, all specimens are scraped against one another to see which makes a deeper scratch (a deeper scratch indicates a softer specimen). On this basis the specimens are ranked 1 (softest) to 16 (hardest) with the results given below. Should we reject the hypothesis that hardness is unaffected by temperature?

| rank | temperature |
|------|-------------|
| 1    | H           |
| 2    | L           |
| 3    | H           |
| 4    | H           |
| 5    | H           |
| 6    | L           |
| 7    | H           |
| 8    | L           |
| 9    | L           |
| 10   | H           |
| 11   | H           |
| 12   | L           |
| 13   | L           |
| 14   | L           |
| 15   | L           |
| 16   | L           |

```{r, echo = FALSE, message = FALSE, warning = FALSE}
metal_temp <- data.frame(
  "temp" = c("H","L","H","H","H","L","H","L","L","H","H","L","L","L","L","L"),
  "rank" = 1:16
)

metal_temp_wrs <- wilcox.test(x = metal_temp$rank[metal_temp$temp == "H"],
                              y = metal_temp$rank[metal_temp$temp == "L"],
                              alternative = "two.sided", exact = TRUE,
                              paired = FALSE)
```

## SEVEN

*(ANSM 5.9)*

Hotpot stoves use a standard oven insulation. To test its effectiveness they take random samples from the production line and heat the ovens selected to $400^{0}C$, noting the time taken to cool to $350^{0}C$, after switching off. For a sample of 8 ovens the times in minutes are: $$15.7,\ 14.8,\ 14.2,\ 16.1,\ 15.3,\ 13.9,\ 17.2,\ 14.9$$ They decide to explore a cheaper insulation, and using this on a sample of 9 the times taken for the same temperature drop are: $$13.7,\ 14.1,\ 14.7,\ 15.4,\ 15.6,\ 14.4,\ 12.9,\ 15.1,\ 14.0$$ Are the firm justified in asserting there is no real evidence of a different rate of heat loss? Obtain a 95 per cent confidence limit for the difference in median heat loss with and without a normality assumption. Comment critically on any differences between your conclusions

```{r, echo = FALSE, message = FALSE, warning = FALSE}
hotpot <- data.frame(
  "insulation" = c(rep("1", 8), rep("2", 9)),
  "minutes" = c(15.7,14.8,14.2,16.1,15.3,13.9,17.2,14.9,
                13.7,14.1,14.7,15.4,15.6,14.4,12.9,15.1,14)
)

hotpot_wrs <- wilcox.test(x = hotpot$minutes[hotpot$insulation == "1"],
                          y = hotpot$minutes[hotpot$insulation == "2"],
                          paired = FALSE, alternative = "two.sided",
                          exact = FALSE, conf.int = TRUE)
```

## EIGHT

*(ANSM 5.14)*

A psychologist notes total time (in seconds) needed to perform a series of simple manual tasks for each of eight children with learning difficulties and seven children without learning difficulties. The times are:

| without difficulties | with difficulties |
|----------------------|-------------------|
| 204                  | 243               |
| 218                  | 228               |
| 197                  | 261               |
| 183                  | 202               |
| 227                  | 343               |
| 233                  | 242               |
| 191                  | 220               |
|                      | 239               |

Use a Smirnov test to find whether the psychologist is justified in asserting these samples are likely to be from different populations.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
diff <- data.frame(
  "grp" = c(rep("without", 7), rep("with", 8)),
  "time" = c(204,218,197,183,227,233,191,
             243,228,261,202,343,242,220,239)
)

diff_ks <- ks.test(
  x = diff$time[diff$grp == "with"],
  y = diff$time[diff$grp == "without"], 
  alternative = "two.sided", exact = TRUE)
```

## NINE

*(ANSM 5.18)*

Carter and Hubert (1985) give data for percentage variation in blood sugar over 1-hour periods for rabbits given two different dose levels of a drug. Is there evidence of a response difference between levels?

| dose 1 | dose 2 |
|--------|--------|
| 0.21   | 1.59   |
| -16.20 | 2.66   |
| -10.10 | -6.27  |
| -8.67  | -2.32  |
| -11.13 | -10.87 |
| 1.96   | 7.23   |
| -10.19 | -3.76  |
| -15.87 | 3.02   |
| -12.81 | 15.01  |

```{r, echo = FALSE, message = FALSE, warning = FALSE}
dose <- data.frame(
  "dose" = c(rep("I", 9), rep("II", 9)),
  "pct_var"= c(0.21,-16.20,-10.10,-8.67,-11.13,1.96,-10.19,-15.87,-12.81,
               1.59,2.66,-6.27,-2.32,-10.87,7.23,-3.76,3.02,15.01)
)
```

## TEN

*(ANSM 5.22)*

The following data are DMF scores for `34` male and `54` female first-year dental students. The DMF score is the total of the numbers of decayed + missing + filled teeth.

| gender | DMF |
|--------|-----|
| male   | 8   |
| male   | 6   |
| male   | 4   |
| male   | 2   |
| male   | 10  |
| male   | 5   |
| male   | 6   |
| male   | 6   |
| male   | 19  |
| male   | 4   |
| male   | 10  |
| male   | 4   |
| male   | 10  |
| male   | 12  |
| male   | 7   |
| male   | 2   |
| male   | 5   |
| male   | 1   |
| male   | 8   |
| male   | 2   |
| male   | 0   |
| male   | 7   |
| male   | 6   |
| male   | 4   |
| male   | 4   |
| male   | 11  |
| male   | 2   |
| male   | 16  |
| male   | 8   |
| male   | 7   |
| male   | 8   |
| male   | 4   |
| male   | 0   |
| male   | 2   |
| female | 4   |
| female | 7   |
| female | 13  |
| female | 8   |
| female | 8   |
| female | 4   |
| female | 14  |
| female | 5   |
| female | 6   |
| female | 4   |
| female | 12  |
| female | 9   |
| female | 9   |
| female | 9   |
| female | 8   |
| female | 12  |
| female | 4   |
| female | 8   |
| female | 8   |
| female | 4   |
| female | 11  |
| female | 6   |
| female | 15  |
| female | 9   |
| female | 8   |
| female | 14  |
| female | 9   |
| female | 8   |
| female | 9   |
| female | 7   |
| female | 12  |
| female | 11  |
| female | 7   |
| female | 4   |
| female | 10  |
| female | 7   |
| female | 8   |
| female | 8   |
| female | 7   |
| female | 9   |
| female | 10  |
| female | 16  |
| female | 14  |
| female | 15  |
| female | 10  |
| female | 4   |
| female | 6   |
| female | 3   |
| female | 9   |
| female | 3   |
| female | 10  |
| female | 3   |
| female | 8   |

```{r, echo = FALSE, message = FALSE, warning = FALSE}
dental <- data.frame(
"grp" = c(rep("m",34), rep("f",54)),
"score" = c(8,6,4,2,10,5,6,6,19,4,10,4,10,12,7,2,5,1,8,2,0,7,6,4,4,11,2,16,8,7,8,4,0,2,
4,7,13,4,8,8,4,14,5,6,4,12,9,9,9,8,12,4,8,8,4,11,6,15,9,8,14,9,8,9,7,12,11,7,4,10,7,8,8,7,9,10,16,14,15,10,4,6,3,9,3,10,3,8)
)
```

Use an asymptotic WMW test to determine whether the DMF score differs significantly between males and females.

```{r,echo = TRUE, message = FALSE, warning = FALSE}
dental_wrs <- wilcox.test(
  x = dental$score[dental$grp == "m"],
  y = dental$score[dental$grp == "f"], paired = FALSE,
  exact = FALSE, alternative = "two.sided")
```

------------------------------------------------------------------------

# 3. Comparing more than two related samples

## ONE

A graduate student performed a pilot study for his dissertation. He wanted to examine the effects of animal companionship on elderly males. He selected 10 male participants from a nursing home. Then he used an ABAB research design, where A represented a week with the absence of a cat and B represented a week with the presence of a cat. At the end of each week, he administered a 20-point survey to measure the quality of life satisfaction. The survey results are presented in `Table 5.9`

| participant | week 1 | week 2 | week 3 | week 4 |
|-------------|--------|--------|--------|--------|
| 1           | 7      | 6      | 8      | 9      |
| 2           | 9      | 8      | 10     | 1      |
| 3           | 15     | 18     | 16     | 17     |
| 4           | 7      | 6      | 8      | 9      |
| 5           | 7      | 8      | 10     | 11     |
| 6           | 10     | 14     | 13     | 11     |
| 7           | 12     | 19     | 11     | 13     |
| 8           | 7      | 4      | 2      | 5      |
| 9           | 8      | 7      | 9      | 5      |
| 10          | 12     | 16     | 14     | 15     |

: Table 5.9

```{r, echo = FALSE, message = FALSE, warning = FALSE}
buddy <- data.frame(
  "participant" = rep(1:10, each = 4),
  "week" = rep(1:4, times = 10),
  "score" = c(7,6,8,9, 9,8,10,1, 15,18,16,17, 7,6,8,9, 7,8,10,11, 
              10,14,13,11, 12,19,11,13, 7,4,2,5, 8,7,9,5, 12,16,14,15)
)
```

Use a **Friedman test** to determine if one or more of the groups are significantly different. Since this is a pilot study, use $\alpha = 0.10$. If a significant difference exists, use **Wilcoxon signed-rank tests** to identify which groups are significantly different. Use the **Bonferroni procedure** to limit the type I error rate. Report your findings

```{r, echo = FALSE, message = FALSE, warning = FALSE}
buddy_fr <- friedman.test(y = buddy$score, 
                          groups = buddy$week, 
                          blocks = buddy$participant)
```

## TWO

A physical education teacher conducted an action research project to examine a strength and conditioning program. Using 12 male participants, she measures the number of curl ups they could do in 1 minute. She measured their performance before the program. Then, she measured their performance at 1 month intervals. `Table 5.10` presents the performance results.

| participant | baseline | month 1 | month 2 |
|-------------|----------|---------|---------|
| 1           | 66       | 67      | 69      |
| 2           | 49       | 50      | 56      |
| 3           | 51       | 52      | 49      |
| 4           | 65       | 65      | 69      |
| 5           | 42       | 43      | 46      |
| 6           | 38       | 39      | 40      |
| 7           | 33       | 31      | 39      |
| 8           | 41       | 41      | 44      |
| 9           | 46       | 47      | 48      |
| 10          | 45       | 46      | 46      |
| 11          | 36       | 33      | 34      |
| 12          | 51       | 55      | 67      |

: Table 5.10: Number of curl ups in a minute

```{r, echo = FALSE, message = FALSE, warning = FALSE}
curlups <- data.frame(
  "participant" = rep(1:12, each = 3),
  "month" = rep(c("baseline", "month1", "month2"), times = 12),
  "score" = c(66,67,69,49,50,56,51,52,49,65,65,69,42,43,46,38,39,40,33,31,39,41,41,44,46,47,48,45,46,46,36,33,34,51,55,67)
)
```

Use a **Friedman test** with $\alpha = 0.05$ to determine if one or more of the groups are significantly different. The teacher is expecting performance gains, so if a significant difference exists, use **one-tailed Wilcoxon signed-rank tests** to identify which groups are significantly different. Use the **Bonferroni** procedure to limit the type I error rate. Report your findings.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
curlups_fr <- friedman.test(y = curlups$score, 
                            groups = curlups$month, 
                            blocks = curlups$participant)

# wilcox.test(x = curlups$score[curlups$month == "baseline"],
#             y = curlups$score[curlups$month == "month1"], alternative = "greater",
#             paired = TRUE, conf.level = 0.05)
# 
# wilcox.test(x = curlups$score[curlups$month == "baseline"],
#             y = curlups$score[curlups$month == "month2"], alternative = "greater",
#             paired = TRUE, conf.level = 0.05)
# 
# wilcox.test(x = curlups$score[curlups$month == "month1"],
#             y = curlups$score[curlups$month == "month2"], alternative = "greater",
#             paired = TRUE, conf.level = 0.05)
```

## THREE

*(ANSM 6.8)*

At the beginning of a session, 12 names are read out in random order to 10 students. Four are names of prominent sporting personalities (Group A), four of national and international politicians (Group B) and four of local dignitaries (Group C). At the end of the session students are asked to recall as many of the names as possible. The numbers recalled were:

| student | group A | group B | group C |
|---------|---------|---------|---------|
| I       | 3       | 2       | 0       |
| II      | 1       | 1       | 0       |
| III     | 2       | 3       | 1       |
| IV      | 4       | 3       | 2       |
| V       | 3       | 2       | 2       |
| VI      | 1       | 0       | 0       |
| VII     | 3       | 2       | 4       |
| VIII    | 3       | 2       | 1       |
| IX      | 2       | 2       | 0       |
| X       | 4       | 3       | 2       |

```{r, echo = FALSE, message = FALSE, warning = FALSE}
remember <- data.frame(
  "student" = rep(as.roman(1:10), each = 3) |> as.character(),
  "grp" = rep(LETTERS[1:3], times = 10),
  "count" = c(3,2,0,1,1,0,2,3,1,4,3,2,3,2,2,
              1,0,0,3,2,4,3,2,1,2,2,0,4,3,2)
)
```

Rank the data within each block (student) and use a Friedman test to assess
evidence of a difference between recall rates for the three groups. In
particular, is the recall rate for group B and/or group C significantly lower
than that for group A? Carry out an ANOVA on the given data. Do the
conclusions agree with the Friedman test? If not, why not?

```{r, echo = FALSE, message = FALSE, warning = FALSE}
remember_fr <- friedman.test(y = remember$count, groups = remember$grp,
                             blocks = remember$student)
```

## FOUR

*(ANSM 6.12)*

Four share tipsters are each asked to predict on 10 randomly selected days
whether the London FTSE Index (commonly known as Footsie) will rise or
fall on the following day. If they predict correctly this is scored as 1, if
incorrectly as 0. Do the scores below indicate differences in tipsters’
ability to predict accurately?

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ftse <- data.frame(
  "day" = rep(1:10, each = 4),
  "tipster" = rep(1:4, times = 10),
  "outcome" = c(1L,1L,1L,1L,0L,1L,1L,1L,0L,1L,0L,0L,1L,1L,1L,0L,1L,0L,1L,0L,1L,1L,1L,1L,1L,1L,1L,1L,0L,0L,1L,1L,1L,0L,0L,0L,1L,0L,1L,1L)
  );ftse
```

| day | tipster 1 | tipster 2 | tipster 3 | tipster 4 |
|-----|-----------|-----------|-----------|-----------|
| 1   | 1         | 1         | 1         | 1         |
| 2   | 0         | 1         | 1         | 1         |
| 3   | 0         | 1         | 0         | 0         |
| 4   | 1         | 1         | 1         | 0         |
| 5   | 1         | 0         | 1         | 0         |
| 6   | 1         | 1         | 1         | 1         |
| 7   | 1         | 1         | 1         | 1         |
| 8   | 0         | 0         | 1         | 1         |
| 9   | 1         | 0         | 0         | 0         |
| 10  | 1         | 0         | 1         | 1         |

## FIVE

*(ANSM 6.14)*

Berry (1987) gives the following data for numbers of premature ventricular
contractions per hour for 12 patients with cardiac arrhythmias when each
is treated with 3 drugs A, B, C

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ventricle <- data.frame(
"patient" = rep(1:12, each = 3),
"drug" = rep(LETTERS[1:3], times = 12),
"count" = c(170,7,0,19,1.4,6,187,205,18,10,0.3,1,216,0.2,22,49,33,30,7,37,3,474,9,5,0.4,0.6,0,1.4,63,36,27,145,26,29,0,0))
```

| patient | drug A | drug B | drug C |
|---------|--------|--------|--------|
| 1       | 170    | 7      | 0      |
| 2       | 19     | 1.4    | 6      |
| 3       | 187    | 205    | 18     |
| 4       | 10     | 0.3    | 1      |
| 5       | 216    | 0.2    | 22     |
| 6       | 49     | 33     | 30     |
| 7       | 7      | 37     | 3      |
| 8       | 474    | 9      | 5      |
| 9       | 0.4    | 0.6    | 0      |
| 10      | 1.4    | 63     | 36     |
| 11      | 27     | 145    | 26     |
| 12      | 29     | 0      | 0      |

Use a Friedman test to investigate differences in response between drugs.
In particular, is there evidence of a difference between drug A and drug B?

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ventricle_fr <- friedman.test(count~drug|patient, data = ventricle)
```

------------------------------------------------------------------------

# 4. Comparing more than two unrelated samples

## ONE

A researcher conducted a study with *n = 15* participants to investigate strength gains from exercise. The participants were divided into three groups and given one of three treatments. Participants' strength gains were measured and ranked. The rankings are presented in `Table 6.8` below

| I   | II  | III |
|-----|-----|-----|
| 7   | 13  | 12  |
| 2   | 1   | 5   |
| 4   | 7   | 16  |
| 11  | 8   | 9   |
| 15  | 3   | 14  |

: Table 6.8

```{r, echo = FALSE, message = FALSE, warning = FALSE}
strength <- data.frame(
  "treatment" = rep(c("I","II","III"), each = 5),
  "str" = c(7,2,4,11,15, 13,1,7,8,3, 12,5,16,9,14)
)
```

Use a **Kruskal-Wallis H test** with $\alpha = 0.05$ to determine if one or more of the groups were significantly different. If a significant difference exists, use a **two-tailed Mann-Whitney U test** or a **two-sample Kolmogorov-Smirnov test** to identify which groups are significantly different. Use the **Bonferroni** procedure to limit the type I error rate. Report your findings

```{r, echo = FALSE, message = FALSE, warning = FALSE}
str_kw <- kruskal.test(x = strength$str, g = strength$treatment)
```

## TWO

A researcher investigated how physical attraction influences the perception among others of a person's effectiveness with difficult tasks. The photographs of 24 people were shown to a focus group. The group was asked to classify the photos into three groups: *very attractive, average* and *very unattractive*. Then, the group ranked the photographs according to their impressions of how capable they were of solving difficult problems. `Table 6.9` shows the classification and rankings of the people in the photos(1 = *most effective*, 24 = *least effective*)

| very attractive | average | very unattractive |
|-----------------|---------|-------------------|
| 1               | 3       | 11                |
| 2               | 4       | 15                |
| 5               | 8       | 16                |
| 6               | 9       | 18                |
| 7               | 13      | 20                |
| 10              | 14      | 21                |
| 12              | 19      | 23                |
| 17              | 22      | 24                |

: Table 6.9

```{r, echo = FALSE, message = FALSE, warning = FALSE}
attraction <- data.frame(
  "group" = rep(c("very_attractive", "average", "very_unattractive"), each = 8),
  "ranking" = c(1,2,5,6,7,10,12,17,
                3,4,8,9,13,14,19,22,
                11,15,16,18,20,21,23,24)
)
```

Use a **Kruskal-Wallis H test** with $\alpha = 0.05$ to determine if one or more of the groups are significantly different. If a significant difference exists, use **two-tailed Mann-Whitney U tests** to identify which groups are significantly different. Use the **Bonferroni** procedure to limit the type I error rate. Report you findings.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
attraction_kw <- kruskal.test(x = attraction$ranking, g = attraction$group)
```

------------------------------------------------------------------------

# 5. Tests for Nominal scale data

## ONE
