---
title: "Non-parametric Methods - Problems"
author: "Gift Pamba"
date: "April-May 2026"
output:
  word_document:
  html_document:
    toc: true
    theme: darkly
---

```{r setup, include=FALSE}
library(tidyverse)
library(DescTools)
library(kableExtra)

```

# Comparing two related samples

## ONE

A teacher wished to determine if providing a bilingual dictionary to students with limited English proficiency improves math test scores. A small class of students **(n = 10)** was selected. Students were given two math tests. Each test covered the same type of math content; however, students were provided a bilingual dictionary on the second test. The data in `Table 3.10` represent the students’ performance on each math test.

| student | Math test without a billingual dictionary | Math test with a billingual dictionary |
|----------|------------------------|------------------------|
| 1 | 30 | 39 |
| 2 | 56 | 46 |
| 3 | 48 | 37 |
| 4 | 47 | 44 |
| 5 | 43 | 32 |
| 6 | 45 | 39 |
| 7 | 36 | 41 |
| 8 | 44 | 40 |
| 9 | 44 | 38 |
| 10 | 40 | 46 |

: Table 3.10

Use a **one-tailed Wilcoxon signed rank** test and a **one-tailed sign test** to determine which testing condition resulted in higher scores. Use $\alpha = 0.05$. Report your findings

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tab1 <- data.frame(
  "student"=seq(1,10,1),
  "before"=c(30,56,48,47,43,45,36,44,44,40),
  "after"=c(39,46,37,44,32,39,41,40,38,46)
)
```

### a

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dictionary <- wilcox.test(x = tab1$before, y = tab1$after,
                          alternative = "greater",
                          paired = TRUE)
```

### b

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dictionary2 <- SignTest(x = tab1$before, y = tab1$after,
                        alternative = "greater")
```

## TWO

A research study was done to investigate the influence of being alone at night on the human male heart rate.Two mean were sent into a wooded area, one at a time, at night, for 20 minutes. They had a heart monitor to record their pulse rate. The second night, the same men were sent into a similar wooded area accompanied by a companion. Thei pulse rate was recorded again. The researcher wanted to see if having a companion would change their pulse rate. The median pulse rates are recorded in `Table 3.11`. Use a **two-tailed Wilcoxon signed-rank test** and a **two-tailed signed test** to determine which condition produced a higher pulse rate. Use $\alpha = 0.05$. Report your findings

| participant | median rate alone | median rate with companion |
|-------------|-------------------|----------------------------|
| A           | 88                | 72                         |
| B           | 77                | 74                         |
| C           | 91                | 80                         |
| D           | 70                | 77                         |
| E           | 80                | 71                         |
| F           | 85                | 83                         |
| G           | 90                | 80                         |
| H           | 82                | 91                         |
| I           | 93                | 86                         |
| J           | 75                | 69                         |

: Table 3.11

```{r, echo=FALSE, message=FALSE, warning=FALSE}
pulse <- data.frame(
  "participant" = LETTERS[1:10],
  "without" = c(88,77,91,70,80,85,90,82,93,75),
  "with" = c(72,74,80,77,71,83,80,91,86,69))
```

### a

```{r, echo=FALSE, message=FALSE, warning=FALSE}
hr <- wilcox.test(x = pulse$without, y = pulse$with,
                  alternative = "two.sided", 
                  paired = TRUE)
```

### b

```{r, echo=FALSE, message=FALSE, warning=FALSE}
hr2 <- SignTest(x = pulse$without, y = pulse$with,
                  alternative = "two.sided")
```

## THREE

A researcher conducts a pilot study to compare two treatments to help obese female teenagers lose weight. She tests each individual in two different treatment conditions. The data in `Table 3.12` provides the number of pounds that each participant lost.

|                  |                 |                 |
|------------------|-----------------|-----------------|
| **participant**  | **treatment 1** | **treatment 2** |
| 1                | 10              | 18              |
| 2                | 20              | 12              |
| 3                | 15              | 16              |
| 4                | 9               | 7               |
| 5                | 18              | 21              |
| 6                | 11              | 17              |
| 7                | 6               | 13              |
| 8                | 12              | 14              |

: Table 3.12

Use a **two-tailed Wilcoxon signed-rank test** and a **two-tailed sign test** to determine which treatment resulted in greater weight loss. Use $\alpha = 0.05$. Report your findings 


```{r, echo = FALSE, message = FALSE, warning = FALSE}
wl <- data.frame(
  "participant" = 1:8,
  "t1" = c(10,20,15,9,18,11,6,12),
  "t2" = c(18,12,16,7,21,17,13,14))
```


### a

```{r, echo = FALSE, message = FALSE, warning = FALSE}
wl_wsr <- wilcox.test(x = wl$t1, y = wl$t2,
                      alternative = "two.sided", paired = TRUE)
```


### b

```{r, echo = FALSE, message = FALSE, warning = FALSE}
wl_sign <- SignTest(x = wl$t1, y = wl$t2,
                    alternative = "two.sided")
```

## FOUR 

Twenty participants in an exercise program were measured on the number of sit-ups they could do before physical exercise (first count) and the number they could do after they had done at least 45 minutes of other physical exercise.(second count) `Table 3.13` shows the results for 20 participants obbtained during two separate physical exercise sessions. Determine the **Effect Size** for a calculated z-score

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ex <- data.frame("participant" = seq(1,10,1),
                 "first" = c(18,19,20,29,15,22,21,30,22,11),
                 "second" = c(20,21,21,20,18,27,24,13,10,10))
# Sample size
n_ex <- nrow(ex)

# Wilcoxon signed-rank hypothesis test for paired samples
ex_wsr <- wilcox.test(x = ex$first, y = ex$second,
                      alternative = "two.sided",
                      paired = TRUE)

# Test statistic
ex_wsr_t <- ex_wsr$statistic

# Normal approximation
## E[T]
ex_wsr_tmean <- (n_ex*n_ex+1)/4

## sd[T]
ex_wsr_tsd <- sqrt((n_ex*(n_ex+1)*(2*n_ex+1))/24)

## z-variate
z_ex <- ((ex_wsr_t-ex_wsr_tmean)/ex_wsr_tsd)

## Effect size
es_ex <- abs(z_ex)/sqrt(n_ex)

cat(paste0("Effect size: ", round(es_ex, 4)))
```


## FIVE

A school is trying to get more students to participate in activities that will make learning more desirable. `Table 3.14` shows the number of activities that each of the 10 students in one class participated in last year before a new activity program was implemented and this year after it was implemented. 
Construct a 95% confidence interval based on the **Wilcoxon signed-rank test** to determine whether the new activity program had a significant positive effect on the student participation

```{r, echo=FALSE, message=FALSE, warning=FALSE}
activity <- data.frame(
  "participant" = seq(1,10,1),
  "lastyr" = c(18,22,10,25,16,14,21,13,28,12),
  "thisyr" = c(20,28,18,23,20,21,17,18,22,21)
)

# Wilcoxon signed-rank hypothesis test
act_wsr <- wilcox.test(x = activity$lastyr, y = activity$thisyr,
                       alternative = "two.sided", paired = TRUE);act_wsr$data.name
```








# Comparing two unrelated samples

## ONE
